---
title: "3-Step ML Auxiliary Variable Integration Using `MplusAutomation`"
subtitle: "*Adding Covariate and Distal Outcome Variables to Mixture Models*"
author: "Adam Garber"
date: '`r format(Sys.time(), "%B, %d,  %Y")`'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE, message = FALSE, warning = FALSE)
```

> This `R` tutorial automates the 3-step ML axiliary variable procedure using the `MplusAutomation` package (Hallquist & Wiley, 2018) to estimate models and extract relevant parameters. To learn more about auxiliary variable integration methods and why multi-step methods are necessary for producing un-biased estimates see Asparouhov & Muthén (2014). 

> The motivation for writing this tutorial is that conducting the 3-step manually is highly error prone as it requires pulling logit values estimated in the step-1 model and adding them in the model statement of the step-2 model (i.e., lots of copying & pasting). In contrast, this approach is fully replicable and provides clear documentation which translates to more reliable research. 

# --------------------------------------------------------------------------------------

$\color{purple}{\text{If using this tutorial to produce tables for publication it would be greatly appreciated if you cite this resource}}$
$\color{purple}{\text{using the citation provided here:}}$

&nbsp;

Garber, A. C. (2021). 3-Step ML Auxiliary Variable Integration Using MplusAutomation. Retrieved from [$\color{blue}{\text{psyarxiv.com/phtxa}}$](psyarxiv.com/phtxa)

# --------------------------------------------------------------------------------------

Associated Github repository here: 

$\color{blue}{\text{https://github.com/garberadamc/3step-ML-Method}}$

# --------------------------------------------------------------------------------------

Load packages
```{r}
library(MplusAutomation) # Conduit between R & Mplus
library(glue)            # Pasting R code into strings
library(here)            # Location, location, location
```

\newpage

# --------------------------------------------------------------------------------------

## "Manual 3-Step" ML Auxiliary Variable Integration Method

# --------------------------------------------------------------------------------------


### Step 1 - Estimate the unconditional model with all covariate & distal outcome variables mentioned in the `auxiliary` statement.

**NOTE**: In this example, Mplus input and output files are directed to the sub-folder `3step_mplus`. Due to the fact that adding auxiliary variables is conducted after enumeration, generally other sub-folders will exist in the top-most `Rproject` folder such as `enum_mplus`, `data`, and `figures`.

```{r}
m_step1  <- mplusObject(
  TITLE = "Step1  (MANUAL 3-STEP ML APPROACH)", 
  VARIABLE = 
   "categorical = X1 X2 X3 X5 X6; 
   
    usevar = X1 X2 X3 X5 X6;
    
    classes = c(3); 
    
    !!! All auxiliary variables to be considered in the final model should be listed here !!!
    auxiliary = 
    COVAR1 DISTAL1 DISTAL2;",
  
  ANALYSIS = 
   "estimator = mlr; 
    type = mixture; 
    starts = 500 100;",
  
  SAVEDATA = 
   "!!! This saved dataset will contain class probabilities and modal assignment columns !!!
    File=3step_savedata_012020.dat;
    Save=cprob;
    Missflag= 999;",
  
  MODEL = "",
  OUTPUT = "",
  
  PLOT = 
    "type = plot3; 
    series = X1 X2 X3 X5 X6(*);",
  
  usevariables = colnames(example_data),
  rdata = example_data)

m_step1_fit <- mplusModeler(m_step1,
                 dataout=here("3step_mplus", "Step1_3step.dat"),
                 modelout=here("3step_mplus", "Step1_3step.inp") ,
                 check=TRUE, run = TRUE, hashfilename = FALSE)
```

# --------------------------------------------------------------------------------------

### Step 2 - Extract logits & saved data from the step 1 unconditional model.

Extract logits for the classification probabilities for the most likely latent class 
```{r}

logit_cprobs <- as.data.frame(m_step1_fit[["results"]]
                                         [["class_counts"]]
                                         [["logitProbs.mostLikely"]])
```

Extract saved data from the step 1 model `mplusObject` named "m_step1_fit"
```{r}

savedata <- as.data.frame(m_step1_fit[["results"]]
                                     [["savedata"]])
```

Rename the column in savedata for "C" and change to "N"
```{r}

colnames(savedata)[colnames(savedata)=="C"] <- "N"

```

### Step 3 (part 1) - Estimate the unconditional model with logits from step 2. 

This model is estimated to check that the class proportions are approximately the same as in step 1. 
```{r}

m_step2  <- mplusObject(
  TITLE = "Step2  (MANUAL 3-STEP ML APPROACH)", 
  
  VARIABLE = 
 "nominal=N;
  USEVAR = n;
  missing are all (999); 
  classes = c(3); ",
  
  ANALYSIS = 
 "estimator = mlr; 
  type = mixture; 
  starts = 0;",
  
  MODEL = 
    glue(
 "%C#1%
  [n#1@{logit_cprobs[1,1]}];
  [n#2@{logit_cprobs[1,2]}];
  
  %C#2%
  [n#1@{logit_cprobs[2,1]}];
  [n#2@{logit_cprobs[2,2]}];
  
  %C#3%
  [n#1@{logit_cprobs[3,1]}];
  [n#2@{logit_cprobs[3,2]}];"),
  
  OUTPUT = "!tech11  tech14 res;",
 
  PLOT = 
 "!type = plot3; 
  !series = X1 X2 X3 X5 X6(*);",
 
  usevariables = colnames(savedata), 
  rdata = savedata)

m_step2_fit <- mplusModeler(m_step2, 
                 dataout=here("3step_mplus", "Step2_3step.dat"), 
                 modelout=here("3step_mplus", "Step2_3step.inp"), 
                 check=TRUE, run = TRUE, hashfilename = FALSE)
```

# --------------------------------------------------------------------------------------

### Step 3 (part 2) - Add covariates & distal outcomes to the model. 

##  Estimate the final SEM Model - Moderation Example

# --------------------------------------------------------------------------------------

**Specification details:**

- This example contains two distal outcomes (`DISTAL1` & `DISTAL2`) and one binary covariate (`COVAR1`).
- Under each class-specific statement (e.g., `%C#1%`) the distal outcomes are mentioned to estimate the intercept parameters. 
- Moderation is specified by mentioning the `"outcome ON covariate;"` syntax under each of the class-specific statements. 
- Note that the binary covariate is centered so that reported distal means (intercepts) are estimated at the weighted average of `COVAR1`.

```{r}

m_step3  <- mplusObject(
  TITLE = "Step3  (MANUAL 3-STEP ML APPROACH)", 
  
  VARIABLE = 
 "nominal = N;
  usevar = n;
  missing are all (999); 
  
  usevar = COVAR1 DISTAL1 DISTAL2;
  classes = c(3); ",
 
  DEFINE =
 "Center COVAR1 (Grandmean);",
  
  ANALYSIS = 
 "estimator = mlr; 
  type = mixture; 
  starts = 0;",
  
  MODEL =
  glue(
 "!!! OUTCOMES = DISTAL1 DISTAL2 !!! 
  !!! MODERATOR = COVAR1         !!! 
  
  %OVERALL%
  DISTAL1 on COVAR1;
  DISTAL1;
  
  DISTAL2 on COVAR1;
  DISTAL2;

  %C#1%
  [n#1@{logit_cprobs[1,1]}];
  [n#2@{logit_cprobs[1,2]}];
  
  [DISTAL1](m01);
  DISTAL1;                    !!! estimate conditional intercept !!!
  DISTAL1 on COVAR1 (s01);    !!! estimate conditional regression !!!

  [DISTAL2] (m1);
  DISTAL2;
  DISTAL2 on COVAR1 (s1);

  %C#2%
  [n#1@{logit_cprobs[2,1]}];
  [n#2@{logit_cprobs[2,2]}];
  
  [DISTAL1](m02);
  DISTAL1;
  DISTAL1 on COVAR1 (s02);

  [DISTAL2] (m2);
  DISTAL2;
  DISTAL2 on COVAR1 (s2);
  
  %C#3%
  [n#1@{logit_cprobs[3,1]}];
  [n#2@{logit_cprobs[3,2]}];
  
  [DISTAL1](m03);
  DISTAL1;
  DISTAL1 on COVAR1 (s03);

  [DISTAL2] (m3);
  DISTAL2;
  DISTAL2 on COVAR1 (s3);"),
  
  MODELCONSTRAINT = 
 "New (diff12 diff13
  diff23 slope12 slope13
  slope23 ndiff12 ndiff13
  ndiff23 nslope12 nslope13
  nslope23);

  diff12 = m1-m2;   ndiff12 = m01-m02;
  diff13 = m1-m3;   ndiff13 = m01-m03;
  diff23 = m2-m3;   ndiff23 = m02-m03;
  slope12 = s1-s2;  nslope12 = s01-s02;
  slope13 = s1-s3;  nslope13 = s01-s03;
  slope23 = s2-s3;  nslope23 = s02-s03;",
  
  MODELTEST =
  ## NOTE: Only a single Wald test can be conducted per model run. Therefore, 
  ## this example requires running separate models for each omnibus test (e.g.,
  ## 4 models; 2 outcomes and 2 slope coefficients). This can be done by
  ## commenting out all but one test and then estimating multiple versions of the model.
   
 "m1=m2;       !!! Distal outcome omnibus Wald test for `DISTAL2` !!!
  m2=m3;       

  !s1=s2;      !!! Slope difference omnibus Wald test `DISTAL2 on COVAR1` !!!
  !s2=s3;

  !m01=m02;    !!! Distal outcome omnibus Wald test for `DISTAL1` !!!
  !m02=m03;

  !s01=s02;   !!! Slope difference omnibus Wald test for `DISTAL2 on COVAR1` !!!
  !s02=s03;",
 
  usevariables = colnames(savedata), 
  rdata = savedata)

m_step3_fit <- mplusModeler(m_step3,
                 dataout=here("3step_mplus", "Step3_3step.dat"), 
                 modelout=here("3step_mplus", "Step3_3step.inp"), 
                 check=TRUE, run = TRUE, hashfilename = FALSE)
```

**End of 3-Step Procedure**

# --------------------------------------------------------------------------------------


### References: 

Asparouhov, T., & Muthén, B. O. (2014). Auxiliary variables in mixture modeling: Three-step approaches using Mplus. Structural Equation Mod- eling, 21, 329–341. http://dx.doi.org/10.1080/ 10705511.2014.915181

Hallquist, Michael N., and Joshua F. Wiley. 2018. “MplusAutomation: An R Package for FacilitatingLarge-Scale Latent Variable Analyses in Mplus.” Structural Equation Modeling, 1–18. https://doi.org/10.1080/10705511.2017.1402334.

Müller, Kirill. 2017.Here:  A Simpler Way to Find Your Files. https://CRAN.R-project.org/package=here.

Muthen L.K., & Muthen B.O. (1998-2017) Mplus User's Guide. Eight Edition. Los Angelos, CA: Muthen & Muthen.

R Core Team. 2019.R: A Language and Environment for Statistical Computing. Vienna, Austria: R Foundation for Statistical Computing. https://www.R-project.org/.

